A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-11 tex/MercuryNumerical.tex

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={I. Hammer, C. Hanhart, C. Körber and C. Müller},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} (\url{https://www.python.org}) as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\footnote{\url{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
A useful choice for this problem is given e.g. by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{eqnarray}

@@

As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
	Of course for very large values of $\alpha$ this linear relation will eventually break down.
	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
}.
The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.

@@

Instead of providing the template to the students, they could build the program from scratch.
Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see~\url{http://www.vpython.org/contents/docs/VPython_Intro.pdf}) or the like.
Also, they could independently research the problem relevant parameters.
\item \textbf{Why can we work in one plane?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{alphadin}
\bibliography{References}

@@

For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} [see also \url{http://vpython.org}].

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All Linux based system come with an installation of \python{}.
Thus, if not already at the newest version, one just has to update existing installations.
This is best done by using a distribution manager.
Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
For this we refer to \url{https://www.macports.org/install.php} or \url{https://docs.brew.sh/Installation.html}.

@@

As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
Following the link \url{https://conda.io/docs/user-guide/install/download.html}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
Both are equally sufficient.
One can install further updates after the initial installation.
In the case of \texttt{Miniconda} (\url{https://conda.io/miniconda.html}) one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.

>>>>

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={C. Körber, I. Hammer, , J.-L. Wynen, C. Müller and C. Hanhart},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\cite{MercuryFactSheet}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
A useful choice for this problem is given e.g. by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{eqnarray}

@@

As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
	Of course for very large values of $\alpha$ this linear relation will eventually break down.
	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
}.
The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
This can be done either by hand, \python{} (e.g., with \texttt{matplotlib} \cite{Matplotlib}), or using another program like, e.g., Excel.
For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.

@@

Instead of providing the template to the students, they could build the program from scratch.
Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see \cite{VPythonIntro}) or the like.
Also, they could independently research the problem relevant parameters.
\item \textbf{Why can we work in one plane?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{unsrt}%{alphadin}
\bibliography{References}

@@

For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} (see also \cite{VPython}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All Linux based system come with an installation of \python{}.
Thus, if not already at the newest version, one just has to update existing installations.
This is best done by using a distribution manager.
Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
For this we refer to \cite{MacPorts} or \cite{Homebrew}.

@@

As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
Following the link \cite{conda}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
Both are equally sufficient.
One can install further updates after the initial installation.
In the case of \texttt{Miniconda} \cite{Miniconda} one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.
